\documentclass[conference]{IEEEtran}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{multirow}
\usepackage{url}
\usepackage{cite}

\hypersetup{
  colorlinks=true,
  linkcolor=black,
  citecolor=black,
  urlcolor=blue
}

\begin{document}

\title{Explainable Chronos: A Framework for Verbalization and Interactive Querying of Probabilistic Time-Series Forecasts}

\author{
\IEEEauthorblockN{[Author~1],
[Author~2],
[Author~3],
Farnood Sotoudeh,
Emanuele Romito}
\IEEEauthorblockA{
Course: 01VIXSM --- Deep Natural Language Processing, A.Y.\ 2025--2026\\
\textit{Repository:} \url{https://github.com/Voyagers-time-series-forecasting/explainable-chronos}}
}

\maketitle

% =====================================================================
\begin{abstract}
We present \textit{Explainable Chronos}, a modular NLP framework built on top of Chronos-2 that makes probabilistic time-series forecasts interpretable and interactively queryable. Extension~1 generates natural-language summaries of quantile forecasts, grounded in extracted numerical features and covariate attributions via Attention Rollout, and scores each sentence for factual consistency using NLI. Extension~2 adds a conversational dialogue interface through which users can query and counterfactually modify forecasts in plain English; a two-tier intent parser (deterministic rule-based with optional LLM fallback) classifies queries into five intent types, applies input modifications, re-runs Chronos-2, and returns a new verbalized response with state preserved across turns. On a 10-query blind evaluation set, Extension~2 achieves 100\% intent accuracy, 100\% task completion rate, mean NLI consistency of 0.9649, and 100\% delta faithfulness across all verifiable cases.
\end{abstract}

\begin{IEEEkeywords}
time-series forecasting, natural language generation, explainability, NLI consistency, dialogue systems, intent parsing, Chronos-2, covariate attribution
\end{IEEEkeywords}


% =====================================================================
\section{Introduction}
% =====================================================================

Probabilistic time-series forecasting models such as Chronos-2~\cite{chronos2} produce quantile predictions that encode rich information about expected trends, uncertainty, and tail risks. However, these numerical outputs are difficult for non-expert stakeholders to interpret: a vector of quantile values communicates nothing intuitively about \emph{why} a forecast was generated, how confident the model is, or what role each input covariate played in shaping the prediction.
This gap between model output and human understanding is precisely what our work is trying to fill.

\vspace{2mm}

Our project addresses this gap by building a modular NLP framework that sits entirely on top of Chronos-2 at inference time and produces two complementary capabilities: (i) automatic natural-language explanations of forecasts grounded in extracted numerical features and covariate attributions (Extension 1), and (ii) an interactive dialogue interface through which users can query and counterfactually modify those forecasts using plain English (Extension 2).

\vspace{2mm}

Extension 1 addresses two complementary challenges: generating natural-language summaries that faithfully describe the content of quantile forecasts, and automatically verifying whether the generated text is consistent with the underlying numerical evidence.
We frame the verification task as a Natural Language Inference (NLI) problem~\cite{bowman2015snli,williams2018mnli}, where a structured rendering of the numerical features serves as the \emph{premise} and each verbalized sentence serves as the \emph{hypothesis}.

% =====================================================================
\section{Problem Statement}
% =====================================================================

\subsection{Extension 1}

\textbf{Input.}
A historical time series $\mathbf{x} = (x_1, \dots, x_T) \in \mathbb{R}^T$ with a matrix of covariates $\mathbf{Z} \in \mathbb{R}^{T \times C}$ with $C$ named covariates.

\vspace{2mm}

\textbf{Task.}
Given a Chronos-2 quantile forecast $\hat{\mathbf{y}}^{(q)}$ for quantiles $q \in \{0.10, 0.50, 0.90\}$ over a horizon $H$:
%
\begin{equation}
  \hat{\mathbf{y}}^{(q)} = \bigl(\hat{y}^{(q)}_1, \dots, \hat{y}^{(q)}_H\bigr), \quad q \in \{0.10, 0.50, 0.90\},
\end{equation}
%
produce a natural-language summary $\mathcal{S} = (s_1, \dots, s_K)$ such that each sentence $s_k$ is factually grounded in a subset of numerical features derived from the quantiles, and then score each sentence for factual consistency via NLI.

\vspace{2mm}

\textbf{Output.}
A verbalization result comprising:
(a) the paragraph-level summary $\mathcal{S}$;
(b) per-sentence grounding metadata mapping each $s_k$ to the numerical features that generated it;
(c) a covariate attribution result containing top-$k$ importance scores derived via Attention Rollout~\cite{abnar2020attentionflow} and per-covariate temporal saliency over the history window;
(d) a consistency report containing per-sentence NLI entailment scores and an aggregate mean score indicating whether the summary is factually consistent;
(e) a semantic similarity score measuring how closely the LLM output aligns with the deterministic template draft, quantifying semantic drift from the structured reference.

\subsection{Extension 2}

\textbf{Input.}
A natural-language query $q$ issued by a user, together with the current dialogue state comprising a covariate set $\mathbf{Z} \in \mathbb{R}^{T \times C}$ and a forecast horizon $H$.

\vspace{2mm}

\textbf{Task.}
Classify $q$ into one of five intent types $\mathcal{I} = \{$\texttt{remove\_covariate}, \texttt{scale\_covariate}, \texttt{change\_horizon}, \texttt{confidence\_query}, \texttt{counterfactual}$\}$, extract any relevant slots (target covariate, scale factor, new horizon), apply the corresponding modification to the forecast inputs, re-run Chronos-2, and generate a natural-language response via the Extension~1 verbalization pipeline. State is preserved across turns.

\vspace{2mm}

\textbf{Output.}
A dialogue response comprising: (a) the classified intent with confidence source; (b) a human-readable description of the modification applied; (c) a new natural-language forecast summary; (d) the delta P50 (change in median forecast relative to the unmodified baseline); (e) an NLI consistency score for the generated response.

% =====================================================================
\section{Methodology}
% =====================================================================

\subsection{Extension 1}

The pipeline consists of four sequential stages: forecast and feature extraction (1), covariate attribution (2), verbalization (3), and NLI consistency scoring (4).

\subsubsection{Forecast and Feature Extraction}

From the Chronos-2 quantile output (P10, P50, P90) we derive seven interpretable feature groups using NumPy and SciPy:

\begin{itemize}
  \item \textbf{Trend}: direction and magnitude of the median forecast.
  \item \textbf{Uncertainty}: level and evolution of the prediction interval width.
  \item \textbf{Asymmetry}: whether the interval skews above or below the median.
  \item \textbf{Tail risks}: binary flags for significant downside or upside extremes.
  \item \textbf{Regime shift}: statistical detection of a mean-level break at the forecast midpoint.
  \item \textbf{Threshold breaches}: step-level violations of user-configurable domain thresholds.
  \item \textbf{Trajectory narration}: segmentation of the P50 curve into monotone phases with turning points.
\end{itemize}




% ----- 4.2 -----
\subsubsection{Covariate Attribution}

The pipeline computes covariate attribution scores via Attention Rollout~\cite{abnar2020attentionflow} and passes the top-$k$ results (default $k=5$) to the verbalizer.

\textbf{Method.}
~\cite{abnar2020attentionflow} observes that raw per-layer attention weights become near-uniform in deep transformers, as each layer's output is a mixture of all previous representations.
Attention Rollout addresses this by composing attention matrices across layers: at each layer the raw weights are averaged over heads, mixed with an identity term to account for residual connections ($A = 0.5 \cdot W_{\text{att}} + 0.5 \cdot I$), and left-multiplied into the accumulated product ($\tilde{A}(l_i) = A(l_i) \cdot \tilde{A}(l_{i-1})$).
The final matrix $\tilde{A}$ approximates how information from each input position propagates to the output.

\textbf{Adaptation to Chronos-2.}
Chronos-2 alternates two attention axes per transformer block: \emph{group attention}, which attends across the target and covariate series at each patch position, and \emph{time attention}, which attends across patches within each series independently.
For \emph{covariate importance}, rollout is applied to the group-attention layers; the per-patch matrices are averaged before composition, and row~0 of $\tilde{A}$ gives the target's attribution to each covariate.
For \emph{temporal saliency}, we ask a complementary question: \emph{which history time steps did each covariate attend to most?}
The same rollout formula is applied independently per covariate to the time-attention layers, producing a normalised distribution over history patches that is interpolated back to the original time steps.
The result is a per-covariate saliency curve, from which we extract the peak step (most attended history moment) and the focus breadth (normalised entropy, measuring how concentrated or diffuse the attention is).


% ----- 4.3 -----
\subsubsection{Verbalization}

We implement two complementary approaches.

\textbf{Template Verbalizer.}
The verbalizer produces a summary in two layers.
The \emph{base layer} always generates one sentence per active feature group: a trend sentence, a trajectory sentence, an uncertainty sentence and conditional risk sentences.
The \emph{rhetorical layer} then adds RST-structured sentences~\cite{mann1988rst} conditioned on feature combinations: \textsc{Concession} (rising trend with downside risk), \textsc{Contrast} (non-flat trend with high uncertainty), \textsc{Elaboration} (regime shift with $p$-value), and \textsc{Cause} (top covariate attribution $>20\%$).
Each sentence is paired with a grounding dictionary recording the numerical features that generated it.

\textbf{LLM Verbalizer.}
The template output serves as a refinement draft for a local LLM and is stored in the verbalization result as \texttt{draft\_summary}.
The prompt is structured as few-shot in-context learning: three worked DRAFT$\to$REWRITE examples are prepended, followed by the actual input consisting of (i)~numerical features, (ii)~grounding triples of the form (\emph{subject}, \emph{predicate}, \emph{object}), (iii)~the template draft, and (iv)~constraints forbidding causal claims, domain-specific speculation, and changes to numerical values.
Greedy decoding is used to maximise factual fidelity.
Because the LLM may freely reorganize content across sentences, each output sentence is mapped to a \emph{combined} grounding aggregating all template feature groundings.
The template verbalizer is not evaluated as a standalone output; it functions solely as a structured reference for the LLM and for semantic similarity scoring.

\textbf{Audience Adaptation.}
The LLM verbalizer accepts a \texttt{persona} parameter ($\in \{$\texttt{analyst},\texttt{executive}$\}$, default \texttt{analyst}) that selects a distinct system prompt and set of few-shot examples while keeping every numerical fact identical.
The \emph{analyst} persona targets quantitative professionals: it preserves technical terminology (P10/P50/P90, prediction interval, regime shift) and produces dense, precise prose.
The \emph{executive} persona targets non-technical decision-makers: the system prompt replaces jargon with plain equivalents (``forecast range'', ``notable change in pattern'', ``lower/central/upper estimate'') and the few-shot rewrites demonstrate a simpler, action-oriented register.
Both personas share the same loaded model weights via \texttt{share\_model\_from()}, adding zero extra GPU memory cost.
Because the grounding triples and NLI premises are identical for both personas, any drop in NLI consistency in the executive output is attributable to genuine factual deviation rather than vocabulary mismatch.



% ----- 4.4 -----
\subsubsection{NLI Consistency Scoring}

For each sentence $s_k$ in the verbalization, a \emph{premise} $\pi_k$ is rendered from its grounding dictionary. A NLI model is used to estimate:
%
\begin{equation}
  P(\text{entail} \mid \pi_k,\, s_k).
\end{equation}
%
The overall consistency score for summary $\mathcal{S}$ is the arithmetic mean over all $K$ sentences:
%
\begin{equation}
  \text{Score}(\mathcal{S}) = \frac{1}{K} \sum_{k=1}^{K} P(\text{entail} \mid \pi_k,\, s_k).
\end{equation}
%
A summary is deemed \emph{consistent} when $\text{Score}(\mathcal{S}) \geq 0.70$.

Premise quality is critical for reliable NLI scoring.
Three enrichments are applied to ensure that each rendered premise contains exactly the facts that the corresponding sentence asserts:
\begin{itemize}
  \item \textbf{Trajectory:} the premise enumerates specific turning-point values and step positions (e.g.\ ``the series peaks near $v$ around step $t$''), matching the level of detail in the hypothesis sentence.
  \item \textbf{Attribution:} the premise includes the covariate's actual effect direction (\emph{positive} or \emph{negative}) alongside the percentage contribution, so that directional claims in the sentence are grounded rather than hallucinated.
  \item \textbf{Temporal focus:} the premise includes the peak attention step index and the named history segment (e.g.\ ``final 15\% of history'') that the sentence cites.
\end{itemize}


\subsection{Extension 2}

Extension~2 builds directly on Extension~1, exposing the verbalization pipeline through a stateful conversational interface. The architecture consists of five components.

\subsubsection{Intent Parser}

The parser uses a two-tier architecture. \textbf{Tier~1} is a deterministic rule-based baseline (\texttt{rule\_parser.py}) consisting of a regex catalog with explicit priority ordering: \texttt{confidence\_query} $\to$ \texttt{remove\_covariate} $\to$ \texttt{scale\_covariate} $\to$ \texttt{change\_horizon} $\to$ \texttt{counterfactual}. Slot extraction handles covariate name resolution via exact match, token overlap, and substring fallback; scale factors via keyword lookup (\textit{double}, \textit{halve}, \textit{triple}) and percentage patterns; and horizon values via digit and word-number patterns (e.g., ``three days'' $\to$ 72 steps for hourly data). \textbf{Tier~2} is an optional LLM fallback (Qwen1.5-1.8B-Chat) that activates only when the rule tier returns \texttt{unknown} and \texttt{use\_llm\_fallback=True} is set; it is disabled by default to keep evaluation deterministic.

\subsubsection{Input Modifier}

A stateless module that applies a \texttt{ParsedIntent} to the current covariate set and horizon, producing modified copies without side effects. For \texttt{remove\_covariate}, the target channel is zeroed out; for \texttt{scale\_covariate}, values are multiplied by the extracted factor; for \texttt{change\_horizon}, the horizon integer is updated. Fuzzy covariate matching (token overlap) handles aliases and partial names.

\subsubsection{Dialogue State Management}

The \texttt{DialogueSystem} maintains state across turns through \texttt{self.covariates} and \texttt{self.horizon}. Every modifying query persists the updated inputs before returning a response, so subsequent turns --- including confidence queries --- operate on the already-modified forecast context.

\subsubsection{Response Generation}

For modifying intents, the modified inputs are passed to the Extension~1 \texttt{VerbalizationPipeline} (with \texttt{enable\_attention=True}), which re-runs Chronos-2, extracts features, computes covariate attribution via Attention Rollout, and generates a new NL summary. The response includes the delta P50 (mean, min, max change relative to the original baseline). For \texttt{confidence\_query}, the system answers directly from the current forecast's P10/P50/P90 without re-running the model. For \texttt{counterfactual}, the system declines to modify historical inputs and redirects the user toward forward-looking alternatives.



% =====================================================================
\section{Evaluation Protocol}
% =====================================================================

\subsection{Extension 1}

Our protocol separates two orthogonal questions: (1)~\emph{Is the forecast accurate?} and (2)~\emph{Is the verbalization a faithful and complete description of what the model predicted?}

% ----- 5.1 -----
\subsubsection{Forecast Accuracy}

Evaluated against observed actuals, independently of the verbalization pipeline.

\begin{itemize}
  \item \textbf{MASE}: ratio of P50 MAE to a na\"ive one-step baseline (history differences);
  \item \textbf{Fair MASE}: ratio of P50 MAE to a multi-step naive baseline that repeats the last observed value over the full horizon.
  \item \textbf{Coverage}: fraction of actuals falling inside the P10--P90 interval;
\end{itemize}

% ----- 5.2 -----
\subsubsection{Verbalization Quality}

Only the LLM verbalizer is evaluated in two audience configurations (\emph{analyst} and \emph{executive}); the template runs internally as a structured reference.
Metrics are organised along two axes:

\textit{a) Faithfulness: Does the text accurately describe the forecast?}
\begin{itemize}
  \item \textbf{NLI consistency}: per-sentence entailment probability from the NLI model (BART-large-MNLI); the mean over all $K$ sentences defines the overall score ($\geq 0.70$ = pass). Premises are enriched with specific numerical values matching the hypothesis (turning-point steps, covariate direction, peak attention index) to ensure that entailment failure signals genuine factual deviation rather than premise under-specification. Because both personas receive identical grounding triples and NLI premises, a drop in the executive score directly measures factual distortion introduced by simplification.
\end{itemize}

\textit{b) Semantic drift: Does the LLM output stay close to the template reference?}
\begin{itemize}
  \item \textbf{Semantic similarity vs.\ template}: cosine similarity between SBERT embeddings (\texttt{all-MiniLM-L6-v2}) of the LLM output and the template draft stored in \texttt{draft\_summary}. A score $\geq 0.50$ indicates acceptable semantic alignment. The executive persona is expected to produce a lower score than the analyst persona, reflecting heavier paraphrasing --- this is the intended behaviour and not a quality failure, provided NLI remains $\geq 0.70$.
\end{itemize}


\subsection{Extension 2}

Extension~2 is evaluated as a dialogue system across two modes. \textit{Intent-only} mode tests the parser without running Chronos-2, reporting intent accuracy and Task Completion Rate (TCR). \textit{Full-pipeline} mode runs Chronos-2 and NLI, additionally reporting mean NLI consistency ($\geq 0.70$ = pass) and delta faithfulness (whether the response correctly describes the direction of the P50 change). Multi-turn state persistence is verified by running sequences of turns and checking that modifications from earlier turns remain in effect. Three query sets are used: DEV (40), TEST (30), and BLIND (10 --- the reported score).


% =====================================================================
\section{Experiments}
% =====================================================================

\subsection{Extension 1}

% ----- 6.1 -----
\subsubsection{Data}

We evaluate on the Seattle Weather dataset.
The forecast target is daily maximum temperature (\texttt{temp\_max}), with three meteorological covariates: minimum temperature (\texttt{temp\_min}), precipitation (\texttt{precipitation}), and wind speed (\texttt{wind}).

% ----- 6.2 -----
\subsubsection{Experimental Setup}

\textbf{Hardware.} NVIDIA A100 GPU via Google Colab.

\textbf{Software and libraries.} Python~3.10+; \texttt{chronos-forecasting} (Chronos-2-small~\cite{chronos2}); \texttt{transformers} (BART-large-MNLI~\cite{lewis2020bart} for NLI, Qwen2.5-7B-Instruct~\cite{qwen} for LLM verbalization); \texttt{sentence-transformers} (\texttt{all-MiniLM-L6-v2} for semantic similarity scoring).

\textbf{Validation method.}
The evaluation protocol from Section~V is applied: forecast accuracy metrics (MASE, coverage) are computed against observed actuals; verbalization quality metrics (NLI consistency, semantic similarity vs.\ template) are computed against the model's own numerical features for both personas.

\textbf{Windowing.}
We evaluate over 200 scenarios. Each scenario uses a history window of 512 steps and a forecast horizon of 96 steps. Windows are positioned at evenly spaced starting points within the test split.

\textbf{Evaluation.}
Both LLM personas (\emph{analyst}, \emph{executive}) are evaluated on every window. They share the same Qwen2.5-7B-Instruct weights and differ only in system prompt and few-shot examples. The template verbalizer runs internally as the structured reference draft.

% ----- 6.3 -----
\subsubsection{Results}

\paragraph*{a) Forecast Accuracy}
Over 200 evaluation windows, Chronos-2-small achieves a mean MASE of $0.998 \pm 0.184$ and a Fair MASE of $0.425 \pm 0.167$, indicating that the median forecast substantially outperforms a flat na\"ive baseline over the 96-step horizon. The P10--P90 interval achieves $91.4\% \pm 4.6\%$ coverage.

\paragraph*{b) Verbalization Quality}

\textit{Faithfulness and drift.}
Table~\ref{tab:verbalization} reports NLI consistency and semantic similarity for both personas across 200 Weather windows.
The expected pattern is: both personas achieve NLI $\geq 0.70$ (factual preservation holds across registers); the executive persona shows lower semantic similarity to the template, reflecting deliberate paraphrasing into simpler language.

\begin{table}[h]
\centering
\caption{LLM verbalization quality by persona on 200 Weather windows (mean $\pm$ std).}
\label{tab:verbalization}
\begin{tabular}{llcc}
\toprule
Persona & Metric & Mean & Std \\
\midrule
\multirow{3}{*}{Analyst}
  & NLI Consistency         & --- & --- \\
  & Semantic Sim.\ vs.\ Template & --- & --- \\
  & NLI PASS rate           & --- & \\
\midrule
\multirow{3}{*}{Executive}
  & NLI Consistency         & --- & --- \\
  & Semantic Sim.\ vs.\ Template & --- & --- \\
  & NLI PASS rate           & --- & \\
\bottomrule
\end{tabular}
\end{table}

\textit{Note:} Results to be filled after running the updated evaluation pipeline (Section~V-B). The key comparison is whether the executive persona maintains NLI $\geq 0.70$ while achieving meaningfully lower semantic similarity — demonstrating that audience adaptation changes form without compromising factual accuracy.

\textit{Covariate attribution.}
Across all 200 windows, \texttt{temp\_min} is ranked as the top covariate in 199 cases ($38.2\% \pm 1.8\%$ mean relative impact), confirming its dominant influence on the \texttt{temp\_max} forecast.


\subsection{Extension 2}

\subsubsection{Data}

The full-pipeline evaluation uses a 100-step window from the ETTh1 dataset~\cite{zhou2021informer} (hourly electricity transformer temperature) with ten named covariates including \texttt{marketing\_spend}, \texttt{weather\_temperature}, \texttt{holiday\_proximity}, and \texttt{social\_media\_mentions}, among others.

\subsubsection{Experimental Setup}

\textbf{Hardware.} CPU (Intel/AMD); no GPU required for intent-only evaluation. Full-pipeline evaluation runs Chronos-2-small on CPU.

\textbf{Evaluation sets.} Three query sets with strict separation: DEV (40 queries, used during parser development), TEST (30 queries, intermediate validation), and BLIND (10 queries, written after patterns were frozen --- the reported score).

\textbf{Evaluation modes.}
\begin{itemize}
  \item \textit{Intent-only} (\texttt{evaluate}): tests parser accuracy, covariate resolution, horizon extraction, and Task Completion Rate (TCR) without running Chronos-2.
  \item \textit{Full pipeline} (\texttt{evaluate-full}): runs Chronos-2 and NLI for each query; additionally measures NLI consistency and delta faithfulness (whether the response correctly describes the direction of the P50 change).
  \item \textit{Multi-turn}: three two-turn sequences run through the public \texttt{query()} interface to verify that covariate modifications persist correctly across turns.
\end{itemize}

\subsubsection{Results}

\paragraph*{a) Intent Classification}

The rule-based parser achieves 100\% intent accuracy and TCR on all three query sets (DEV: 40 queries, TEST: 30, BLIND: 10). Table~\ref{tab:ext2_intent} reports the per-intent breakdown on the BLIND set.

\begin{table}[h]
\centering
\caption{Intent classification accuracy on BLIND set (10 queries).}
\label{tab:ext2_intent}
\begin{tabular}{lcc}
\toprule
Intent Type & Correct & Accuracy \\
\midrule
\texttt{change\_horizon}   & 2/2 & 100\% \\
\texttt{confidence\_query} & 2/2 & 100\% \\
\texttt{counterfactual}    & 1/1 & 100\% \\
\texttt{remove\_covariate} & 2/2 & 100\% \\
\texttt{scale\_covariate}  & 3/3 & 100\% \\
\midrule
\textbf{Overall}           & \textbf{10/10} & \textbf{100\%} \\
\bottomrule
\end{tabular}
\end{table}

\paragraph*{b) Full Pipeline}

Table~\ref{tab:ext2_full} reports full-pipeline metrics on the BLIND set using real ETTh1 data.

\begin{table}[h]
\centering
\caption{Full-pipeline evaluation on BLIND set (ETTh1, 10 queries).}
\label{tab:ext2_full}
\begin{tabular}{lc}
\toprule
Metric & Value \\
\midrule
Intent accuracy       & 100\% \\
Task Completion Rate  & 100\% \\
NLI consistency (mean) & 0.9649 \\
NLI PASS rate         & 100\% \\
Delta faithfulness    & 100\% (7 verifiable) \\
\bottomrule
\end{tabular}
\end{table}

\paragraph*{c) Multi-Turn State Persistence}

Three two-turn sequences were evaluated: (1) remove covariate then query confidence, (2) scale covariate then change horizon, (3) two successive covariate modifications. All three sequences passed, with 100\% intent accuracy and 100\% state persistence across 2 verifiable checks per run. In an extended stress test, the system executes three successive modifications --- remove \texttt{weather\_temperature}, double \texttt{social\_media\_mentions}, change horizon to 48 steps --- followed by a confidence query. The removed covariate remained removed, the scaled covariate remained doubled, the horizon remained 48, and all tasks completed successfully.

\paragraph*{d) LLM Fallback}

The optional LLM fallback tier was evaluated on five indirect queries that the rule-based parser could not resolve. The rule-only parser achieved 3/5 accuracy; enabling the LLM fallback with structured-output normalisation recovered both unknown cases, achieving 5/5 accuracy at a mean latency of 1.31 seconds per query.

\paragraph*{e) Qualitative Case Study}

To demonstrate end-to-end integration, we walk through a four-turn dialogue using a synthetic scenario with ten covariates over a 14-step horizon.

\textbf{Turn~1 --- Covariate removal.} Query: \textit{``What if there were no marketing spend?''} The system zeros out \texttt{marketing\_spend} and re-runs Chronos-2. The delta P50 is near zero (mean $= +0.0007$), indicating low marginal impact; \texttt{previous\_day\_sales} emerges as the dominant driver (30.4\%) once marketing spend is removed. NLI: 0.9594 (PASS).

\textbf{Turn~2 --- Confidence query.} Query: \textit{``How confident are you in this forecast?''} The system answers directly from the modified forecast's P10/P50/P90 --- no re-run. It reports high and stable uncertainty, with mean interval width 0.43 (61.0\% of forecast level), P10~$= 0.49$, P90~$= 0.91$. NLI: 0.9494 (PASS).

\textbf{Turn~3 --- Horizon change.} Query: \textit{``Show me the next three weeks.''} The parser correctly resolves ``three weeks'' to 504 steps (hourly data). The \texttt{marketing\_spend} zeroing from Turn~1 remains in effect. Delta P50: mean $= -0.0148$. NLI: 0.9438 (PASS).

\textbf{Turn~4 --- Historical counterfactual.} Query: \textit{``What would have happened if sales had been higher last month?''} The system declines to modify historical inputs, identifies \texttt{previous\_day\_sales} as the target, and redirects: \textit{``I can show you what the forecast would look like if `previous\_day\_sales' were removed or scaled in future periods.''} NLI: 0.9494 (PASS).

% =====================================================================
\section{Conclusions and Takeaways}
% =====================================================================

\emph{To be completed after experimental results are available.}


% =====================================================================
% REFERENCES
% =====================================================================
\begin{thebibliography}{7}

\bibitem{chronos2}
A.~F.~Ansari, L.~Stella, C.~Turkmen, et~al.,
``Chronos-2: From Univariate to Universal Forecasting,''
\textit{arXiv preprint arXiv:2510.15821v1}, 2025.

\bibitem{bowman2015snli}
S.~R.~Bowman, G.~Angeli, C.~Potts, and C.~D.~Manning,
``A large annotated corpus for learning natural language inference,''
in \textit{Proc.\ EMNLP}, 2015, pp.~632--642.

\bibitem{williams2018mnli}
A.~Williams, N.~Nangia, and S.~R.~Bowman,
``A broad-coverage challenge corpus for sentence understanding through inference,''
in \textit{Proc.\ NAACL-HLT}, 2018, pp.~1112--1122.

\bibitem{abnar2020attentionflow}
S.~Abnar and W.~Zuidema,
``Quantifying Attention Flow in Transformers,''
in \textit{Proc.\ ACL}, 2020, pp.~4190--4197.

\bibitem{mann1988rst}
W.~C.~Mann and S.~A.~Thompson,
``Rhetorical Structure Theory: Toward a functional theory of text organization,''
\textit{Text}, vol.~8, no.~3, pp.~243--281, 1988.

\bibitem{lewis2020bart}
M.~Lewis, Y.~Liu, N.~Goyal, et~al.,
``BART: Denoising Sequence-to-Sequence Pre-training for Natural Language Generation, Translation, and Comprehension,''
in \textit{Proc.\ ACL}, 2020, pp.~7871--7880.

\bibitem{qwen}
Qwen Team,
``Qwen2.5 Technical Report,''
\textit{arXiv preprint arXiv:2412.15115}, 2024.

\bibitem{zhou2021informer}
H.~Zhou, S.~Zhang, J.~Peng, et~al.,
``Informer: Beyond Efficient Transformer for Long Sequence Time-Series Forecasting,''
in \textit{Proc.\ AAAI}, 2021, pp.~11106--11115.

\end{thebibliography}

% =====================================================================
\appendix
% =====================================================================

\textbf{Extension 1: Illustrative Success and Failure Cases}

\textbf{Success case}
The extracted features indicate a flat trend ($+0.0415$ per step), high and widening uncertainty, both downside and upside risk, and a statistically significant regime shift ($p < 10^{-5}$). The dominant covariate is \texttt{temp\_min} at 39.0\%, followed by \texttt{precipitation} (30.8\%) and \texttt{wind} (30.2\%).
The LLM produces:
\begin{quote}
\emph{``The forecast suggests stability over the next 96 periods, starting from 22.20 and ending near 27.34, which is 23.2\% higher. Prediction intervals are wide and expanding, indicating increasing uncertainty. There is both downside and upside risk, with bounds up to 20\% above or below current levels. A structural break is detected mid-horizon, driven mainly by temperature (39.0\%), followed by precipitation (30.8\%) and wind (30.2\%).''}
\end{quote}
Every numerical claim and directional fact maps directly to a feature in the grounding, and the summary passes the NLI consistency check ($\geq 0.70$).

\textbf{Failure case}
Suppose the LLM had instead produced:
\begin{quote}
\emph{``The forecast indicates a declining trend over the 96-period horizon, likely driven by adverse seasonal weather patterns and temperature inversions. Long-term conditions appear to be deteriorating, suggesting the structural shift observed at the midpoint may persist well beyond the forecast window.''}
\end{quote}
The NLI premise for the first sentence contains only the flat trend classification and a slope of $+0.0415$ --- no mention of ``declining,'' ``seasonal patterns,'' or ``temperature inversions.''
The NLI model assigns an entailment probability near~0.05, correctly flagging the direction reversal (flat vs.\ declining) and the hallucinated domain-specific causal content.
This illustrates the key failure mode of the LLM verbalizer: despite the grounding constraints in the prompt, the model can introduce plausible-sounding but factually unsupported language that the NLI metric reliably catches.


\vspace{4mm}
\textbf{Extension 2: Illustrative Success and Failure Cases}

\textbf{Success case (actionable intent).}
Query: \textit{``What if there were no marketing spend?''}
The parser correctly classifies the intent as \texttt{remove\_covariate}, extracts \texttt{marketing\_spend} as the target, zeros out the covariate channel, and re-runs Chronos-2. The generated response correctly reports that \texttt{previous\_day\_sales} becomes the dominant driver (30.4\%) and that the forecast remains sharply rising. Delta P50: mean $= +0.0007$. NLI consistency: 0.9594 (PASS).

\textbf{Failure case (non-actionable intent).}
Query: \textit{``What would have happened if sales had been higher last month?''}
The system correctly classifies this as \texttt{counterfactual} and declines to execute it, since only future inputs can be modified. Rather than failing silently, it redirects the user: \textit{``I can show you what the forecast would look like if `previous\_day\_sales' were removed or scaled in future periods.''} This illustrates the system's graceful degradation: non-actionable queries are handled with an informative response rather than an error.

\end{document}
